Estimating duration in the presence of misclassification

### 2024-08-02 cis-imperfect-testing.tex
edits to corrections

In this section, I introduce a simple model of false negatives (\ie allowing $\psens < 1$) into the posterior density derived in \cref{E-perf-test:sec:model}; I continue to assume the probability of a false positive is negligible.
Using a simple model, in particular a constant test sensitivity, means calculating the likelihood remains tractable.
Additionally, I limit the number of permutations of false negative and true positive tests that I consider have non-negligible probability.
Specifically, I assume that, if $i(j)$ is tested during infection episode $j$, either: the individual has a single false negative test and the episode ends before their next test, or the first test is a false positive test.
This assumption is reasonable because false negatives normally occur when the viral load is low, which is normally late in the infection episode.
For the same reason, I assume there is at most one false negative test following the final true positive test in the episode.

@@

Introducing false negatives means that $O(j)$, containing the relevant information for episode $j$, is now random, even if $B_j$ and $E_j$ are known.
Therefore, the results between the tests that bound episode $j$'s beginning and end times should now enter into the likelihood.
For tractability, I consider only tests between the negative tests providing an upper bound on the length of episode $j$.

>>>>

In this section, I introduce a simple model of false negatives (\ie allowing $\psens < 1$) into the posterior density derived in \cref{E-perf-test:sec:model}; I continue to assume the probability of a false positive is negligible.
Using a simple model, in particular a constant test sensitivity, means calculating the likelihood remains tractable.
Additionally, I limit the number of permutations of false negative and true positive tests that I consider to have non-negligible probability.
Specifically, I assume that, if $i(j)$ is tested during infection episode $j$, either: the individual has a single false negative test and the episode ends before their next test, or the first test is a true positive test.
This assumption is reasonable because false negatives normally occur when the viral load is low, which is normally late in the infection episode.
For the same reason, I assume there is at most one false negative test following the final true positive test in the episode.

@@

Introducing false negatives means that $O(j)$, containing the test results which would bound the length of episode $j$ in the absence of false negatives, is a random variable even after conditioning on $B_j$ and $E_j$.
Therefore, the results between the tests that bound episode $j$'s beginning and end times should now enter into the likelihood.
For tractability, I consider only tests between the negative tests providing an upper bound on the length of episode $j$.

### 2024-08-02 cis-perfect-testing.tex
Christl p.97 comment
edits to corrections

A related problem is estimating how long insect and bird nests survive for, as detected events are doubly interval censored and some are undetected.
In the context of nest survival, Bayesian methods have been by augmenting the data with the unobserved times of infection~\autocite{heBayesiana,heBayesian,caoModeling}.
The data augmentation approach creates a parameter for each detected infection.
This increases the dimensionality of the problem greatly, making it computationally infeasible with the large sample size of the CIS within the SRS.

@@

The test results for any individual is fully determined by any infection episodes that occur in that individual and their test schedule.
The set of times $i(j)$ tests positive at due to $W_j$ is $\posResults(W_j) = \{ t \in \sched_{i(j)} \ssep B_j \leq t \leq E_j \}$.
The negative test immediately before the start of the episode (if there is one) is $\negResults^{(l)}(W_j) = \max \{ t \in \sched_{i(j)} \ssep t < B_j \}$.
The negative test immediately after the end of the episode (if there is one) is $\negResults^{(r)}(W_j) = \min \{ t \in \sched_{i(j)} \ssep t > E_j \}$.

Next, I define $O(j)$ to contain the test results due to $W_j$ that give information on $W_j$'s length, and hence are relevant to inference about $\vec{\theta}$.
If $\posResults(W_j) = \emptyset$, $\min \posResults(W_j) \notin [1, T]$, or either $\negResults^{(l)}$ or $\negResults^{(r)}$ is undefined then the episode is undetected, in which case $O(j) = \emptyset$.
Otherwise, $O(j) = [l_j^{(b)}, r_j^{(b)}, l_j^{(e)}, r_j^{(e)}, i(j)]^T$, where $i(j)$ is the individual in which the episode occurs; $l_j^{(b)} = \negResults^{(l)}(W_j) + 1$ and $r_j^{(b)} = \min \posResults(W_j)$ are the earliest and latest time the episode could have begun and produced the observed series of test results respectively; and $l_j^{(e)} = \max \posResults(W_j)$ and $r_j^{(e)} = \negResults^{(r)}(W_j) - 1$ are the earliest and latest time the episode could have ended respectively.
These definitions coincide with the way the dataset is constructed (see \cref{E-biology-data:sec:cis-episodes}).

@@

As we assume $O(j) = \vec{\nu}_k$ and $O(j') = \vec{\nu}_{k'}$ are independent for $j \neq j'$:
\begin{align}
  \vec{n} \mid \ntot, \vec{\theta} &\dist \MN(\ntot, \vec{p})
\intertext{that is:}
  p(\vec{n} \mid \ntot, \vec{\theta}) &= \frac{\ntot!}{\nnodet!\prod_{k=1}^{N_E} n_k!} p_u^{\nnodet} \prod_{k=1}^{N_E} p_k^{n_k}.
  \label{perf-test:eq:multinomial-ll}
\end{align}
\begin{figure}
\makebox[\textwidth][c]{\includegraphics[width=0.9\paperwidth]{cis-perfect-testing/regions_diag}}
\thisfloatpagestyle{empty}
\caption[Episode regions]{%
  Each dot is a combination of $b_j$ and $e_j$ for an arbitrary individual $i$.
  Each box, bounded by dashed lines, are combinations giving $O(j) = \vec{\nu}_k$, $k$ corresponding to the numeric label; labels are arbitrary but unique across all individuals.
  $i$ had negative tests at times 0, 7, 14, 56, and 84 (not shown) and positive tests at times 21 and 28.
  The purple region corresponds to the doubly interval censored episode in this individual.
  That is, $n_8 = 1$ and $n_k = 0$ for $k = 1, \dots, 7, 9, \dots, 15$.
  The red region corresponds to combinations giving $O(j) = \emptyset$.
  Impossible region violates $b_j \leq e_j$.
}
\label{perf-test:fig:partitionSpace}
\end{figure}

@@

Recall the definitions $p_k = \prob(O(j) = \vec{\nu}_k \mid \vec{\theta})$ where $O(j)$ is the random variable containing all relevant information for episode $j$ and $\vec{\nu}_k = [l^{(b)}_k, r^{(b)}_k, l^{(e)}_k, r^{(e)}_k, i_k]^T$ is an arbitrary element from $O(j)$'s state space, excluding undetected infections.

@@

Uninformative Beta priors on the hazard are problematic in this context.
An uninformative prior would be of the form $\lambda_t \dist \BetaDist(\alpha, \alpha)$.
Commonly, $\alpha$ is chosen to be $0.5$ or $1$ (a uniform prior).
The former, $\alpha = 0.5$, is Jefferys’ prior~\autocite{jeffreysInvariant}, commonly considered an uninformative prior because it is invariant to parameterization~\autocite[53]{gelmanBDA}.
Even though these priors are uninformative on the hazard, they become highly informative on the survival. 
Specifically, they tend to favour shorter survival times (see \cref{perf-test:fig:flat-prior}).
An intuitive explanation of this follows from the fact that priors of this form have an expected value of 0.5.
The expected survival time, $\E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right)$, is therefore equal to $0.5^{t-1}$, a quantity that declines rapidly.
This prior expresses extreme scepticism in estimates from previous studies, such as \cref{E-ATACCC} and the meta-analysis of \textcite{cevikShedding}.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics{cis-perfect-testing/flat-prior}}
  \caption[Uninformative priors for the hazard]{
    (A) Density of Beta(1, 1) flat prior and Beta(0.5, 0.5) uninformative priors.
    (B) Prior predictive survival time (mean and 95\% CrI).
    The implied prior on the survival time is short when using either prior on the hazard at each time.
  }
 \label{perf-test:fig:flat-prior}
\end{figure}

@@

I first approximate the \cref{E-ATACCC} posterior estimate of the hazard as $\logit{\vec{h}} \dist \MNorm(\vec{\mu}_A, \matr{\Sigma}_A)$ where $\vec{h} = [h_1, \dots, h_{\dmax}]^T$ is the hazard, and $\vec{\mu}_A$ and $\matr{\Sigma}_A$ are the mean and covariance matrix computed from the samples of $\logit{\vec{h}}$ derived using the method of \cref{E-ATACCC:sec:duration}.
Therefore, $\vec{h}$ represents a sample from the approximation of \cref{E-ATACCC}'s posterior estimate of the hazard before increasing the uncertainty to account for the strong model assumptions.
Using a multivariate normal, as opposed to multiple univariate distribution for each $h_t$, preserves the correlation between the hazards.
% The logit transformation maps from the $[0, 1]$ interval to the full real line.
The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC-hazard} and \cref{perf-test:fig:approximate-ATACCC-survival}).
\begin{figure}
  \thisfloatpagestyle{empty}
  % \vspace{-1cm}
  \makebox[\textwidth][c]{\includegraphics{cis-perfect-testing/ataccc-approximation-hazard}}
  \caption[Approximating the ATACCC posterior hazard as a logit-normal]{Comparison of the posterior estimate of the hazard from \cref{E-ATACCC} and its approximation with a logit-normal distribution (kernel density estimate smoothed) for hazards between day 4 and 15. The density for smaller hazards is concentrated very close to 0. \label{perf-test:fig:approximate-ATACCC-hazard}}
\end{figure}
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics{cis-perfect-testing/ataccc-approximation-survival}}
  \caption[Approximating the ATACCC posterior survival]{Comparison of the posterior estimate of the survival from \cref{E-ATACCC} and its approximation with a logit-normal distribution (median and central 95\% interval). The survival is a function of all the hazards up until that point, therefore the correlation between the hazards must be well-approximated to have these agree. \label{perf-test:fig:approximate-ATACCC-survival}}
\end{figure}

>>>>

A related problem is estimating how long insect and bird nests survive for
In this context detected events are doubly interval censored and some events are undetected; Bayesian methods have been developed which augment the data with the unobserved times of infection~\autocite{heBayesiana,heBayesian,caoModeling}.
The data augmentation approach creates a parameter for each detected infection.
This increases the dimensionality of the problem greatly, making it computationally infeasible with the large sample size of the CIS within the SRS.

@@

The test results for any individual is fully determined by any infection episodes that occur in that individual and their test schedule.
The set of times $i(j)$ tests positive at due to $W_j$ is $\posResults(W_j) = \{ t \in \sched_{i(j)} \ssep B_j \leq t \leq E_j \}$.
The negative test immediately before the start of the episode (if there is one) is $\negResults^{(b)}(W_j) = \max \{ t \in \sched_{i(j)} \ssep t < B_j \}$.
The negative test immediately after the end of the episode (if there is one) is $\negResults^{(e)}(W_j) = \min \{ t \in \sched_{i(j)} \ssep t > E_j \}$.

Next, I define $O(j)$ to contain the test results due to $W_j$ that give information on $W_j$'s length, and hence are relevant to inference about $\vec{\theta}$.
If $\posResults(W_j) = \emptyset$, $\min \posResults(W_j) \notin [1, T]$, or either $\negResults^{(b)}$ or $\negResults^{(e)}$ is undefined then the episode is undetected, in which case $O(j) = \emptyset$.
Otherwise, $O(j) = [l_j^{(b)}, r_j^{(b)}, l_j^{(e)}, r_j^{(e)}, i(j)]^T$, where $i(j)$ is the individual in which the episode occurs; $l_j^{(b)} = \negResults^{(b)}(W_j) + 1$ and $r_j^{(b)} = \min \posResults(W_j)$ are the earliest and latest time the episode could have begun and produced the observed series of test results respectively; and $l_j^{(e)} = \max \posResults(W_j)$ and $r_j^{(e)} = \negResults^{(e)}(W_j) - 1$ are the earliest and latest time the episode could have ended respectively.
These definitions coincide with the way the dataset is constructed (see \cref{E-biology-data:sec:cis-episodes}).

@@

By the independence assumption:
\begin{align}
  \vec{n} \mid \ntot, \vec{\theta} &\dist \MN(\ntot, \vec{p})
\intertext{that is:}
  p(\vec{n} \mid \ntot, \vec{\theta}) &= \frac{\ntot!}{\nnodet!\prod_{k=1}^{N_E} n_k!} p_u^{\nnodet} \prod_{k=1}^{N_E} p_k^{n_k}.
  \label{perf-test:eq:multinomial-ll}
\end{align}
\begin{figure}
\makebox[\textwidth][c]{\includegraphics[width=0.9\paperwidth]{cis-perfect-testing/regions_diag}}
\thisfloatpagestyle{empty}
\caption[Episode regions]{%
  Each dot is a combination of $b_j$ and $e_j$ for an arbitrary individual $i$.
  Each box, bounded by dashed lines, are combinations giving $O(j) = \vec{\nu}_k$, $k$ corresponding to the numeric label; labels are arbitrary but unique across all individuals.
  $i$ had negative tests at times 0, 7, 14, 56, and 84 (not shown) and positive tests at times 21 and 28.
  The purple region corresponds to the doubly interval censored episode in this individual.
  That is, $n_8 = 1$ and $n_k = 0$ for $k = 1, \dots, 7, 9, \dots, 15$.
  The red region corresponds to combinations giving $O(j) = \emptyset$.
  Impossible region violates $b_j \leq e_j$.
}
\label{perf-test:fig:partitionSpace}
\end{figure}

@@

Recall the definitions $p_k = \prob(O(j) = \vec{\nu}_k \mid \vec{\theta})$ where $O(j)$ contains the relevant test results for episode $j$ and $\vec{\nu}_k = [l^{(b)}_k, r^{(b)}_k, l^{(e)}_k, r^{(e)}_k, i_k]^T$ is an arbitrary element from $O(j)$'s state space, excluding undetected infections.

@@

Uninformative Beta priors on the hazard are problematic in this context.
An uninformative prior would be of the form $\lambda_t \dist \BetaDist(\alpha, \alpha)$.
Commonly, $\alpha$ is chosen to be $0.5$ or $1$ (a uniform prior).
The former, $\alpha = 0.5$, is Jefferys' prior~\autocite{jeffreysInvariant}, commonly considered an uninformative prior because it is invariant to the parameterization used~\autocite[53]{gelmanBDA}.
Even though these priors are uninformative on the hazard, they become highly informative on the survival. 
Specifically, they tend to favour shorter survival times (see \cref{perf-test:fig:flat-prior}).
An intuitive explanation of this follows from the fact that priors of this form have an expected value of 0.5.
The expected survival time, $\E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right)$, is therefore equal to $0.5^{t-1}$, a quantity that declines rapidly.
This prior expresses extreme scepticism in estimates from previous studies, such as \cref{E-ATACCC} and the meta-analysis of \textcite{cevikShedding}.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics{cis-perfect-testing/flat-prior}}
  \caption[Uninformative priors for the hazard]{
    (A) Density of Beta(1, 1) flat prior and Beta(0.5, 0.5) uninformative priors.
    (B) Prior predictive survival time (mean and 95\% CrI).
    The implied prior on the survival time is short when using either prior on the hazard at each time.
  }
 \label{perf-test:fig:flat-prior}
\end{figure}

@@

I first approximate the \cref{E-ATACCC} posterior estimate of the hazard as $\logit{\vec{h}} \dist \MNorm(\vec{\mu}_A, \matr{\Sigma}_A)$ where $\vec{h} = [h_1, \dots, h_{\dmax}]^T$ is the hazard, and $\vec{\mu}_A$ and $\matr{\Sigma}_A$ are the mean and covariance matrix computed from the samples of $\logit{\vec{h}}$ derived using the method of \cref{E-ATACCC:sec:duration}.
Therefore, $\vec{h}$ represents a sample from the approximation of \cref{E-ATACCC}'s posterior estimate of the hazard before increasing the uncertainty to account for the strong model assumptions.
Using a multivariate normal, as opposed to an independent univariate distribution for each $h_t$, preserves the correlation between the hazards.
% The logit transformation maps from the $[0, 1]$ interval to the full real line.
The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC-hazard} and \cref{perf-test:fig:approximate-ATACCC-survival}).
\begin{figure}
  \thisfloatpagestyle{empty}
  % \vspace{-1cm}
  \makebox[\textwidth][c]{\includegraphics{cis-perfect-testing/ataccc-approximation-hazard}}
  \caption[Approximating the ATACCC posterior hazard as a logit-normal]{Comparison of the posterior estimate of the hazard from \cref{E-ATACCC} and its approximation with a logit-normal distribution (kernel density estimate smoothed) for hazards between day 4 and 15. The density for smaller hazards is concentrated very close to 0. \label{perf-test:fig:approximate-ATACCC-hazard}}
\end{figure}
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics{cis-perfect-testing/ataccc-approximation-survival}}
  \caption[Approximating the ATACCC posterior survival]{Comparison of the posterior estimate of the survival from \cref{E-ATACCC} and its approximation with a logit-normal distribution (median and central 95\% interval). The survival is a function of all the hazards up until that point, therefore the correlation between the hazards must be well-approximated to have these agree. \label{perf-test:fig:approximate-ATACCC-survival}}
\end{figure}

### 2024-08-02 incidence-prevalence.tex
edits to corrections

\emph{Duration}, or \emph{time-to-event}, distributions can be expressed in several equivalent ways~\autocite[17]{sunStatisticala}.
The duration distribution I will be concerned with is the distribution of the duration of detectability.
Consider an individual $i$ infected on day $B_i$ that is prevalent during the inclusive interval $[B_i, E_i]$ (as in \cref{E-biology-data:sec:cis-episodes}); if $B_i = E_i$ then $i$ is prevalent for exactly one day during which they are detectable.
I define this period as an \emph{infection episode}, beginning on day $B_i$ and ending on day $E_i$; therefore, with a duration $D_i = E_i - B_i + 1$.
I model the $D_i$s as a discrete iid (independent and identically distributed) random variables; extending the methodology to account for covariates is left to further work.
The duration is a positive integer, \ie $D_i \in \{1, 2, \dots, \dmax \}$, where $\dmax$ is the maximum duration.
The distribution of $D_i$ can be expressed through the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cdf (cumulative density function), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
Alternatively, the \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(d - 1)$ with $S(1) = 1$.
%(a common alternative convention for survival analysis with discrete variables is to define the survival function as $\prob(D >d)$).
The \emph{hazard} on day $d$ is the probability of recovering on day $d$ conditional on still being positive on day $d$: $\lambda_{d} = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
Throughout this thesis, the survival function will be the primary quantity of interest, and normally parameterized in terms of the hazard; these two quantities are related by $S(d) = \prod_{i=1}^{d-1} (1 - \lambda_{i})$.

>>>>

\emph{Duration}, or \emph{time-to-event}, distributions can be expressed in several equivalent ways~\autocite[17]{sunStatisticala}.
The duration distribution I will be concerned with is the distribution of the duration of detectability.
Consider an individual $i$ infected on day $B_i$ and is prevalent during the inclusive interval $[B_i, E_i]$ (as in \cref{E-biology-data:sec:cis-episodes}); if $B_i = E_i$ then $i$ is prevalent for exactly one day during which they are detectable.
I define this period as an \emph{infection episode}, beginning on day $B_i$ and ending on day $E_i$; therefore, with a duration $D_i = E_i - B_i + 1$.
I model the $D_i$s as a discrete iid (independent and identically distributed) random variables; extending the methodology to account for covariates is left to further work.
The duration is a positive integer, \ie $D_i \in \{1, 2, \dots, \dmax \}$, where $\dmax$ is the maximum duration.
The distribution of $D_i$ can be expressed through the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cdf (cumulative density function), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
Alternatively, the \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(d - 1)$ with $S(1) = 1$.
%(a common alternative convention for survival analysis with discrete variables is to define the survival function as $\prob(D >d)$).
The \emph{hazard} on day $d$ is the probability of recovering on day $d$ conditional on still being positive on day $d$: $\lambda_{d} = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
Throughout this thesis, the survival function will be the primary quantity of interest, and normally parameterized in terms of the hazard; these two quantities are related by $S(d) = \prod_{i=1}^{d-1} (1 - \lambda_{i})$.

### 2024-08-02 introduction.tex
edits to corrections

On 31 Dec 2019, China informed the WHO (World Health Organization) of cases of pneumonia of unknown cause in Wuhan City, Hubei province~\autocite{whoNovelPneumonia2020}.
By 7 Jan 2020, the cause of the pneumonia had been identified as a novel coronavirus which would go on to be named SARS-CoV-2~\autocite{whoNovelCov2020}.
SARS-CoV-2 spread rapidly around the world: on 31 Jan 2020 the first cases were confirmed in England in two Chinese nationals~\autocite{bbcFirstUKCOVID}.
However, introductions from Spain and France in Mar 2020 were the most important sources of the epidemic in England~\autocite{duplessisEstablishment}.

The evolution of the epidemic in England was characterised by a succession of waves, with the context affected by three major types of changes: the imposition and relaxation of non-pharmaceutical interventions (NPIs), the emergence of new variants of the virus, and the roll-out of vaccination.
There were many NPIs, but the most dramatic were the three national lockdowns imposed on 23 Mar 2020, 5 Nov 2020, and 4 Jan 2021 which legally required everyone to stay at home except for limited exceptions~\autocite{ifgTimeline}.
Variants are genetic changes in the virus, the most consequential of which were the three variants of concern, named Alpha, Delta, and Omicron, that spread widely in England due to their growth advantage; prior to the emergence of Alpha, there were many variants in circulation, collectively known as ``pre-Alpha variants''~\autocite{lythgoeLineage}.
Alpha became the dominant variant in Dec 2020, Delta in May 2021, and Omicron in Dec 2021~\autocite{lythgoeLineage}.
The first vaccination in England was on 8 Dec 2020, with 68\% of adults having received two doses by Jul 2020 and 85\% by Oct 2021; initially, two doses was a full course, but later additional doses were offered, especially following the rise of the Omicron variant~\autocite{naoCovidVaccination}.

@@

The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys in representative samples of the general population.
These were designed to avoid the biases of community testing and observe the general population.
These surveys use RT-PCR (Real-Time Polymerase Chain Reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
The first survey was the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
The CIS was a longitudinal, household-based study, testing a large number of individuals weekly or monthly for SARS-CoV-2 infection (see \cref{E-intro:sec:cis} for a detailed description).

>>>>

On 31 Dec 2019, China notified the WHO (World Health Organization) that they had detected several cases of pneumonia of unknown cause in Wuhan City, Hubei province~\autocite{whoNovelPneumonia2020}.
By 7 Jan 2020, the cause of the pneumonia had been identified as a novel coronavirus which would go on to be named SARS-CoV-2~\autocite{whoNovelCov2020}.
SARS-CoV-2 spread rapidly around the world: on 31 Jan 2020 the first cases were confirmed in England in two Chinese nationals~\autocite{bbcFirstUKCOVID}.
However, introductions from Spain and France in Mar 2020 were the most important sources of the epidemic in England~\autocite{duplessisEstablishment}.

The evolution of the epidemic in England was characterised by a succession of waves, with the context affected by three major types of changes: the imposition and relaxation of NPIs (non-pharmaceutical interventions), the emergence of new variants of the virus, and the roll-out of vaccination.

There were many NPIs, but the most dramatic were the three national lockdowns imposed on 23 Mar 2020, 5 Nov 2020, and 4 Jan 2021 which legally required everyone to stay at home except for limited exceptions~\autocite{ifgTimeline}.

Variants are genetic changes in the virus, with those causing significant challenges known as \emph{variants of concern}~\autocite{whoVOC} and labelled using a Greek letter.
Three variants of concern, Alpha, Delta, and Omicron, successively dominated all other variants in England; prior to the emergence of Alpha, there were many variants in circulation, collectively known as ``pre-Alpha variants''~\autocite{lythgoeLineage}.
Alpha became the dominant variant in Dec 2020, Delta in May 2021, and Omicron in Dec 2021~\autocite{lythgoeLineage}.

The first vaccination in England was on 8 Dec 2020, with 68\% of adults having received two doses by Jul 2020 and 85\% by Oct 2021; initially, two doses was a full course, but later additional doses were offered, especially following the rise of the Omicron variant~\autocite{naoCovidVaccination}.

@@

The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys which aimed to recruit representative samples of the general population; such surveys are considered the gold-standard for conducting population surveillance~\autocite{mccabeCISincidence}.
To achieve this aim, they invited individuals selected at random from the population; however, not all individuals selected responded to the invite introducing some non-representativeness (see \cref{E-biology-data:sec:MRP}).
These designs aim to avoid the biases of community testing and observe the general population.
These surveys use RT-PCR (Real-Time Polymerase Chain Reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
The first survey was the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
The CIS was a longitudinal, household-based study, testing a large number of individuals weekly or monthly for SARS-CoV-2 infection (see \cref{E-intro:sec:cis} for a detailed description).
